---
title: "EpiGIFT_Seurat : Investigation on Variable Features"
author: "Laëtitia Racine"
date: "2023-02-15"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "show"
    toc: true
    toc_float: false
    theme: journal
---
    
```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(sctransform)
library(stringr)
library(tidyverse)
library(kableExtra)

```

```{r, Working directories and external script}

start_time = Sys.time()
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/")
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/"))
directory_output = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/")

```


We have lists from FindMarkers with different thresholds and Wilcoxon test.    
We have two lists from EpiGift depending the number of bin fixed (bin 70 => 130 genes, bin 30 => 365 genes).
We have a certain number of common genes between Epigift lists and FM lists.  
We want to see :  
1) if the overlapping number increase when we change FM threshold on logFC  
2) if this number can be just "luck" so we compare to a random selection of genes  

```{r, Input loading}

# List of 130 genes with threshold bin 70 in EpiGift
epi_130 = read.csv2(paste0(directory, "data/", "20230220_EpiGift_Variable130Genes.csv"), sep = ",")
epi_130 = epi_130[,2]
# List of 365 genes with threshold bin 30 in EpiGift
epi_365 = read.csv2(paste0(directory, "data/", "20230220_EpiGift_Variable365Genes.csv"), sep = ",")
epi_365 = epi_365[,2]
# List of FM genes and object merge_fm
dir = paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/")
list_FM_genes = readRDS(paste0(dir, "list_genes_FM_different_thresholds.rds"))
merge_fm = readRDS(paste0(dir, "merge_fm.rds"))

```
  
```{r}

# See if the overlap increases proportionally or not

list_overlap_epi130 = list()
list_overlap_epi365 = list()
list_overlap_random130 = list()
list_overlap_random365 = list()

for (i in 1:length(list_FM_genes)) {
  
  threshold = names(list_FM_genes)[i]
  
  # List of 130 and 365 random genes extract from the matrix to compare
  random_130 = sample(rownames(merge_fm), 130)
  random_365 = sample(rownames(merge_fm), 365)
  
  # List of FM genes
  genes = list_FM_genes[[i]]
  
  list_overlap_epi130[threshold] = list(intersect(epi_130, genes))
  list_overlap_epi365[threshold] = list(intersect(epi_365, genes))
  list_overlap_random130[threshold] = list(intersect(random_130, genes))
  list_overlap_random365[threshold] = list(intersect(random_365, genes))
  
}

tab_comp = data.frame(Nbgenes_FindMarkers = lengths(list_FM_genes),
                      Nbgenes_overlap_epi130 = lengths(list_overlap_epi130),
                      Nbgenes_overlap_epi365 = lengths(list_overlap_epi365),
                      Nbgenes_overlap_random130 = lengths(list_overlap_random130),
                      Nbgenes_overlap_random365 = lengths(list_overlap_random365)) %>%
  tibble::rownames_to_column("FindMarkers_logfc_threshold")

tab_comp %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

tab_plot = tab_comp %>% 
  tidyr::pivot_longer(cols=c(2:ncol(tab_comp)),
                      names_to="Category",
                      values_to="Nbgenes")

plot = ggplot(tab_plot, aes(x=FindMarkers_logfc_threshold, y=Nbgenes, 
                            shape=Category, col=Category)) +
  geom_point()
plot

ggsave(plot = plot, filename = paste0(directory_output, "overlap_EpiGiftgenes_FindMarkersgenes.svg")) 
write.csv2(x = tab_comp, file = paste0(directory_output, "tab_comp_EpiGift_FindMarkers.csv"))

```


```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results='hide'}

# Clean working space and memory 
rm(list = ls())
gc()

```
