---
title: "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_pvalue"
author: "Laëtitia Racine & Giota Kyratzi"
date: "2023-03-09"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(stringr)
library(ggplot2)
library(ggVennDiagram)
library(kableExtra)

```

```{r, Working directory and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/")
start_time = Sys.time()
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_pvalue/"))
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_pvalue/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_pvalue/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "Epi_FM_functions_constants.R"))

# Define a color code for list category 

color_code = c(
  "common_FM_EG" = "#6922FB",
  "only_EG" = "#F73131",
  "only_FM" = "#2E9EFF"
)

```


<br><br><br>


# Script's objective and reminder of previous analysis

To write = see if pvalue of common gene are smaller than one of non common genes

Abbreviations : 
- FM = FindMarkers() function, classical Seurat analysis based on fold change of average  
- EG = EpiGIFT method, new analysis based on each individual  
  
```{r, Load Input data}

genes_EG_list = read.csv2(paste0(directory_data, "20230309_EpiGift_Variable174Genes_pSNP5gt40.csv"), sep =",")

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/"))
genes_FM_list = readRDS(paste0(dir, "/", "list_219_genes_FindMarkers_th0.15_wilcox_pval0.05.rds"))
genes_FM_list = genes_FM_list[["wilcox"]]

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/"))

# Giota's direcrtory
# genes_FM_list <- readRDS("/shared/ifbstor1/projects/strength_field_u/Epigenetics/FindMarkers_th0.15_pval0.05_genes.rds")
# genes_EG_list <- read.csv("/shared/ifbstor1/projects/strength_field_u/Epigenetics/list_174genes_pSNP5gt40.csv")

common_list = readRDS(paste0(dir, "/", "list_genes_CTRL-DON_commonEGFM.rds"))
only_FM_list = readRDS(paste0(dir, "/", "list_genes_CTRL-DON_onlyFM.rds"))
only_EG_list = readRDS(paste0(dir, "/", "list_genes_CTRL-DON_onlyEG.rds"))

```

```{r}

cat("CTRL versus DON : \n",
    "FindMarkers - Seurat :",  nrow(genes_FM_list), "genes \n",
    "EpiGIFT :", nrow(genes_EG_list), "genes \n",
    "Common genes :", length(common_list), "\n",
    "Only FindMarkers :", length(only_FM_list), "\n",
    "Only EpiGift :", length(only_EG_list), "\n")

# change color to correspond to color code 
ggVennDiagram(x = list(FindMarkers= genes_FM_list$gene, 
                       EpiGift= genes_EG_list$GeneName)) +
  theme(legend.position = "none") 

```

```{r, Add list category information in tables}

genes_EG_list = genes_EG_list %>%
  dplyr::mutate(list_category = case_when(GeneName %in% common_list ~ "common_FM_EG",
                                          GeneName %in% only_EG_list ~ "only_EG",
                                          GeneName %in% only_FM_list ~ "only_FM"))

genes_FM_list = genes_FM_list %>%
  dplyr::mutate(list_category = case_when(gene %in% common_list ~ "common_FM_EG",
                                          gene %in% only_EG_list ~ "only_EG",
                                          gene %in% only_FM_list ~ "only_FM"))

```



<br><br><br>



# pValue distribution of FindMarkers 

```{r}

# Reorder rows by ascending order of pvalue adjusted
genes_FM_list = genes_FM_list %>% dplyr::arrange(p_val_adj)
genes_order = genes_FM_list$gene
genes_FM_list$gene = factor(genes_FM_list$gene , levels = genes_order)

```

```{r, fig.height = 6, fig.width = 20}

# Represent the data
ggplot(data = genes_FM_list, aes(x=gene, y=p_val_adj, col=list_category)) +
  geom_point() +
  scale_color_manual(values = color_code) +
  xlab(label = "Gene Name") +
  ylab(label = "pvalue adjusted") +
  ggtitle(label = "Ordered pvalue distribution for FM list of genes") +
  theme(axis.text.x = element_text(angle = 90, size = 6))

```

```{r, fig.height = 6, fig.width = 20}

# Zoom in
ggplot(data = genes_FM_list, aes(x=gene, y=p_val_adj, col=list_category)) +
  geom_point() +
  scale_color_manual(values = color_code) +
  xlab(label = "Gene Name") +
  ylab(label = "pvalue adjusted") +
  coord_cartesian(ylim= c(0, 0.00025)) +
  scale_y_continuous(breaks=seq(0,0.00025, 0.00001)) +
  ggtitle(label = "Ordered pvalue distribution for FM list of genes",
          subtitle = "Zoom on values near 0") +
  theme(axis.text.x = element_text(angle = 90, size = 6))

```

```{r, fig.height = 10, fig.width = 20}

# Zoom in
ggplot(data = genes_FM_list, aes(x=gene, y=p_val_adj, col=list_category)) +
  geom_point() +
  scale_color_manual(values = color_code) +
  xlab(label = "Gene Name") +
  ylab(label = "pvalue adjusted") +
  coord_cartesian(ylim= c(0, 0.000005)) +
  scale_y_continuous(breaks=seq(0,0.000005, 0.0000001)) +
  ggtitle(label = "Ordered pvalue distribution for FM list of genes",
          subtitle = "Zoom on values near 0") +
  theme(axis.text.x = element_text(angle = 90, size = 6))

```

<br><br><br>



# pSNP5 distribution of EpiGIFT

```{r}
pval5 <- as.data.frame(fread("/shared/ifbstor1/projects/strength_field_u/Epigenetics/CTRL_DON/BinFullData/pval5.csv"))

genes_EG_list$pSNP5 <- pval5[pval5[,1] %in% genes_EG_list$GeneName,2] # logical vector that corresponds to the vector on the left
genesFMpSNP5 <- pval5[pval5[,1] %in% genes_FM_list$gene,]
genes_FM_list$pSNP5 <- genesFMpSNP5[match(genes_FM_list$gene, genesFMpSNP5[,1]),2]
```

```{r}

# Reorder rows by ascending order of pSNP5
genes_EG_list = genes_EG_list %>% dplyr::arrange(pSNP5)
genes_order = genes_EG_list$GeneName
genes_EG_list$GeneName = factor(genes_EG_list$GeneName , levels = genes_order)

```

```{r, fig.height = 6, fig.width = 20}

# Represent the data
ggplot(data = genes_EG_list, aes(x=GeneName, y=pSNP5, col=list_category)) +
  geom_point() +
  scale_color_manual(values = color_code) +
  xlab(label = "Gene Name") +
  ylab(label = "pSNP5") +
  ggtitle(label = "Ordered pSNP5 distribution for EG list of genes") +
  theme(axis.text.x = element_text(angle = 90, size = 6))

```

```{r, fig.height = 6, fig.width = 20}

# Zoom in
ggplot(data = genes_EG_list, aes(x=GeneName, y=pSNP5, col=list_category)) +
  geom_point() +
  scale_color_manual(values = color_code) +
  xlab(label = "Gene Name") +
  ylab(label = "pSNP5") +
  coord_cartesian(ylim= c(0, 230)) +
  scale_y_continuous(breaks=seq(0,0.00025, 0.00001)) +
  ggtitle(label = "Ordered pSNP5 distribution for EG list of genes",
          subtitle = "Zoom on values between 0 and around 200") +
  theme(axis.text.x = element_text(angle = 90, size = 6))

```

<br><br><br>



# pvalue distribution of FindMarkers against pvalue distribution of EpiGIFT

**Step 1 : table with all the genes and FindMarkers pvalue**  
FindMarkers() pvalue : the table genes_FM_list contains only_FM and common_EG_FM genes. We need to add only_EG genes in the table.  
We manually calculated pvalue for those genes in scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes.Rmd.  
We will join the table to have a complete one with all the genes of the three categories.  
NB : for FM method, a good pvalue is really close to 0.  
```{r}

# Genes only EG
dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/"))
genes_EG_FMpval = readRDS(paste0(dir, "/", "EG_FindMarkerscalcul_CTRLDON.rds"))

# Giota's direcrtory
genes_EG_FMpval = readRDS("/shared/ifbstor1/projects/strength_field_u/Epigenetics/CTRL_DON/Git/exp/EG_FindMarkerscalcul_CTRLDON.rds")


genes_EG_FMpval = genes_EG_FMpval %>% 
  dplyr::select(gene, p_val_adj) %>%
  dplyr::mutate(list_category = case_when(gene %in% common_list ~ "common_FM_EG",
                                          gene %in% only_EG_list ~ "only_EG")) 
  
# Genes only FM and common EG FM
genes_FM = genes_FM_list %>% dplyr::select(gene, p_val_adj, list_category)

# Join the table 
tab_all_genes = full_join(x = genes_EG_FMpval %>% dplyr::filter(list_category != "common_FM_EG"), 
                          y = genes_FM, by = c("gene", "p_val_adj", "list_category"))
tab_all_genes = tab_all_genes %>% dplyr::rename("pvalueFM" = "p_val_adj") 

# Display the result
tab_all_genes %>%
  kable() %>%
  kable_styling()


# save intermediate table 

saveRDS(tab_all_genes, paste0(directory_output, "tab_all_genes_pvalueFM.rds"))
```

<br>

**Step2 : Add EG pvalue for all the genes**  
```{r}

# Load EG pSNP5 table
pval5 <- as.data.frame(fread("/shared/ifbstor1/projects/strength_field_u/Epigenetics/CTRL_DON/BinFullData/pval5.csv"))

# Add information in the global table 
genes_pSNP5 <- pval5[pval5[,1] %in% tab_all_genes$gene,]
tab_all_genes$pSNP5 <- genes_pSNP5[match(tab_all_genes$gene, genes_pSNP5[,1]),2]


# Display the result
tab_all_genes %>%
  kable() %>%
  kable_styling()


# save intermediate table 

saveRDS(tab_all_genes, "/shared/ifbstor1/projects/strength_field_u/Epigenetics/CTRL_DON/Git/exp/tab_all_genes_pvalues.rds")
```


<br>

**Step3 : Graphical representation with pSNP5 on y axis and pvalueFM on x axis**  
```{r}

# Create a table with name of the gene, list_category, pvalue in FM, pvalue in EG
# gene list_category (common, onlyEG, onlyFM) pvalueEG pvalueFM
# Draw one graph for each SNP : pvalueEG on y axis, pvalueFM on x axis and different colors for list_category

ggplot(data = tab_all_genes, aes(x=pvalueFM, y=pSNP5, fill=list_category)) +
  geom_point() +
  scale_fill_manual(values = color_code, limits = force)

```


<br>



# Conclusion
   
To write 




<br><br><br>

****

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r}

# Clean working space and memory 
rm(list = ls())
gc()

```
