---
title: "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram"
author: "Laëtitia Racine"
date: "2023-05-25"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}  
div.red {background-color:#FFA996; border-radius: 5px; padding: 8px;}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(stringr)
library(ggplot2)
library(kableExtra)

```

```{r, Working directory and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/")
start_time = Sys.time()
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram/"))
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram/", 
                         current_date))
directory_output = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "Epi_FM_functions_constants.R"))

# Define a color code for list category 
color_code = c(
  "common_FM_EG" = "#6922FB",
  "only_EG" = "#F73131",
  "only_FM" = "#2E9EFF"
)

```

 
 
<br><br><br>



# Overview 

```{r, Load Input data}

genes_EG_list = read.csv2(paste0(directory_data, "20230309_EpiGift_Variable174Genes_pSNP5gt40.csv"), sep =",")

dir1 = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/"))
genes_FM_list = readRDS(paste0(dir1, "/", "list_219_genes_FindMarkers_th0.15_wilcox_pval0.05.rds"))
genes_FM_list = genes_FM_list[["wilcox"]]

# Giota's direcrtory
# genes_FM_list <- readRDS("/shared/ifbstor1/projects/strength_field_u/Epigenetics/FindMarkers_th0.15_pval0.05_genes.rds")
# genes_EG_list <- read.csv("/shared/ifbstor1/projects/strength_field_u/Epigenetics/list_174genes_pSNP5gt40.csv")

dir2 = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/"))
common_list = readRDS(paste0(dir2, "/", "list_genes_CTRL-DON_commonEGFM.rds"))
only_FM_list = readRDS(paste0(dir2, "/", "list_genes_CTRL-DON_onlyFM.rds"))
only_EG_list = readRDS(paste0(dir2, "/", "list_genes_CTRL-DON_onlyEG.rds"))

list_seurat_obj = readRDS(paste0(directory_data, "list_seurat_obj_epi.rds"))

```

Reminder :  
```{r}

cat("CTRL versus DON : \n",
    "FindMarkers - Seurat :",  nrow(genes_FM_list), "genes \n",
    "EpiGIFT :", nrow(genes_EG_list), "genes \n",
    "Common genes :", length(common_list), "\n",
    "Only FindMarkers :", length(only_FM_list), "\n",
    "Only EpiGift :", length(only_EG_list), "\n")

# change color to correspond to color code 
ggVennDiagram(x = list(FindMarkers= genes_FM_list$gene, 
                       EpiGift= genes_EG_list$GeneName)) +
  theme(legend.position = "none") 

```


<br><br><br>



# UMAP with the different lists of genes 

We subset the initial count matrix to keep only the genes of each list and we visualize the cells distribution on the UMAP based on those genes expression.   

```{r, function definition}

# subset the matrix depending on the list of genes 
# PCA
# UMAP

```

## UMAP based on FM list of genes (219 genes)

## UMAP based on FM ONLY list of genes (142 genes)

## UMAP based on EG list of genes (174 genes)

## UMAP based on EG ONLY list of genes (97 genes)

## UMAP based on FM and EG common list of genes (77 genes)


<div class = "red">
Comments 
</div>


<br><br><br>


# Histogram of mRNA number => see QC graphs in classical analysis

## mRNA number per gene per cell

## mRNA number per gene per cell - cumulative sum 

## mRNAnumber 


<div class = "red">
Comments
</div>

<br><br><br>


# Conclusion 



<br><br><br>

****

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r}

# Clean working space and memory 
rm(list = ls())
gc()

```
