---
title: "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram"
author: "Laëtitia Racine"
date: "2023-05-25"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}  
div.red {background-color:#FFA996; border-radius: 5px; padding: 8px;}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(stringr)
library(ggplot2)
library(kableExtra)
library(gridExtra)

```

```{r, Working directory and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/")
start_time = Sys.time()
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram/"))
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram/", 
                         current_date))
directory_output = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_UMAPHistogram/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "Epi_FM_functions_constants.R"))

# Define a color code for list category 
color_code = c(
  "common_FM_EG" = "#6922FB",
  "only_EG" = "#F73131",
  "only_FM" = "#2E9EFF",
  "CTRL" = "#009E73",
  "DON" = "#0072B2" 
)

```

 
 
<br><br><br>



# Overview 

```{r, Load Input data}

genes_EG_list = read.csv2(paste0(directory_data, "20230309_EpiGift_Variable174Genes_pSNP5gt40.csv"), sep =",")

dir1 = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/"))
genes_FM_list = readRDS(paste0(dir1, "/", "list_219_genes_FindMarkers_th0.15_wilcox_pval0.05.rds"))
genes_FM_list = genes_FM_list[["wilcox"]]

# Giota's direcrtory
# genes_FM_list <- readRDS("/shared/ifbstor1/projects/strength_field_u/Epigenetics/FindMarkers_th0.15_pval0.05_genes.rds")
# genes_EG_list <- read.csv("/shared/ifbstor1/projects/strength_field_u/Epigenetics/list_174genes_pSNP5gt40.csv")

dir2 = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/"))
common_list = readRDS(paste0(dir2, "/", "list_genes_CTRL-DON_commonEGFM.rds"))
only_FM_list = readRDS(paste0(dir2, "/", "list_genes_CTRL-DON_onlyFM.rds"))
only_EG_list = readRDS(paste0(dir2, "/", "list_genes_CTRL-DON_onlyEG.rds"))
list_seurat_obj_sct = readRDS(paste0(dir2, "/", "list_seurat_obj_epi_sct.rds"))

```

Reminder :  
```{r}

cat("CTRL versus DON : \n",
    "FindMarkers - Seurat :",  nrow(genes_FM_list), "genes \n",
    "EpiGIFT :", nrow(genes_EG_list), "genes \n",
    "Common genes :", length(common_list), "\n",
    "Only FindMarkers :", length(only_FM_list), "\n",
    "Only EpiGift :", length(only_EG_list), "\n")

# change color to correspond to color code 
ggVennDiagram(x = list(FindMarkers= genes_FM_list$gene, 
                       EpiGift= genes_EG_list$GeneName)) +
  theme(legend.position = "none") 

```


<br><br><br>



# UMAP with the different lists of genes 

We work from the list of seurat object, normalized with SCT transformation and variable.features.rv.th = 1.3.  
As a first step, we select the two lists that correspond to CTRL and DON and we merge the Grange objects. 
Then, we add the variable feature list in the new object.  
```{r}

# Select drugs and merge 
merge_obj = merge(x = list_seurat_obj_sct[["CTRL"]], 
                  y = list_seurat_obj_sct[["DON"]],
                  add.cell.ids = c("CTRL", "DON"),
                  merge.data = TRUE)

# Variable features
VariableFeatures(merge_obj) = rownames(merge_obj[["SCT"]]@scale.data)

# Set Identity to the condition name
merge_obj = SetIdent(object = merge_obj, value = "orig.ident")

```
  
We want to visualize the cell's distribution (cell's separation between drugs) on UMAP depending the list of genes we used to draw it.    
We subset the initial count matrix to keep only the genes of the studied list and we calculate the PCA and the UMAP based on those genes' expression.  
https://github.com/satijalab/seurat/issues/1963  
```{r, functions definition}

umap_highlight = function(seurat_obj, color_code) {

  ind_umap = list()

  for (i in 1:length(unique(seurat_obj$orig.ident))) {

  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")

  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))

  # Plot
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.05,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")

  ind_umap[[i]] = plot_umap

  }

  # function's output
  return(ind_umap)

 }
 

umap_genespecific = function(merge_obj, list_genes, color_code) {
  
  # subset the matrix depending on the list of genes 
  merge_sub = subset(merge_obj, features = list_genes)
  cat("Matrix count has now", nrow(merge_sub), "genes.")
  
  # PCA 
  merge_sub = RunPCA(object = merge_sub, 
                     assay = "SCT", 
                     npcs = 100,
                     reduction.name = "pca", 
                     seed.use = 42, 
                     verbose = FALSE,
                     approx = FALSE)
  # UMAP
  merge_sub = RunUMAP(object = merge_sub, 
                      assay = "SCT", 
                      dims = c(1:40), 
                      reduction =  "pca", 
                      reduction.name = "umap", 
                      verbose = FALSE)
  
  # plot_UMAP
  plot_merged = DimPlot(merge_sub,
                        reduction = "umap",
                        group.by = "orig.ident") +
    scale_color_manual(values = color_code, limits = force)
  
  plot_separated = umap_highlight(merge_sub, 
                                  color_code)
  plot_separated = do.call("grid.arrange", 
                           c(plot_separated, ncol = 2))
  
  # output of the function
  results = list(merge_sub = merge_sub,
                 plot_merged_UMAP = plot_merged,
                 plot_sep_UMAP = plot_separated)
  return(results)

}

```


<br>

## UMAP based on all the genes of the study (22084 genes)

```{r, fig.show="hide"}

all_genes_UMAP = umap_genespecific(merge_obj, rownames(merge_obj), color_code)

```

```{r, fig.width=8, fig.height=10}

all_genes_UMAP$plot_merged_UMAP / all_genes_UMAP$plot_sep_UMAP

```

<br>

## UMAP based on FM list of genes (219 genes)

```{r, fig.show="hide"}

FM_genes_UMAP = umap_genespecific(merge_obj, genes_FM_list$gene, color_code)

```

```{r, fig.width=8, fig.height=10}

FM_genes_UMAP$plot_merged_UMAP / FM_genes_UMAP$plot_sep_UMAP

```

<br>

## UMAP based on FM ONLY list of genes (142 genes)

```{r, fig.show="hide"}

FM_only_genes_UMAP = umap_genespecific(merge_obj, only_FM_list, color_code)

```

```{r, fig.width=8, fig.height=10}

FM_only_genes_UMAP$plot_merged_UMAP / FM_only_genes_UMAP$plot_sep_UMAP

```

```{r, fig.width = 24, fig.height = 60}

FeaturePlot(object = FM_only_genes_UMAP$merge_sub, features = only_FM_list, ncol = 6)

```

<br>

## UMAP based on EG list of genes (174 genes)

```{r, fig.show="hide"}

EG_genes_UMAP = umap_genespecific(merge_obj, genes_EG_list$GeneName, color_code)

```

```{r, fig.width=8, fig.height=10}

EG_genes_UMAP$plot_merged_UMAP / EG_genes_UMAP$plot_sep_UMAP

```

<br>

## UMAP based on EG ONLY list of genes (97 genes)

```{r, fig.show="hide"}

EG_only_genes_UMAP = umap_genespecific(merge_obj, only_EG_list, color_code)

```

```{r, fig.width=8, fig.height=10}

EG_only_genes_UMAP$plot_merged_UMAP / EG_only_genes_UMAP$plot_sep_UMAP

```

```{r, fig.width = 24, fig.height = 50}

FeaturePlot(object = EG_only_genes_UMAP$merge_sub, features = only_EG_list, ncol = 6)

```

<br>

## UMAP based on FM and EG common list of genes (77 genes)

```{r, fig.show="hide"}

FM_EG_com_genes_UMAP = umap_genespecific(merge_obj, common_list, color_code)

```

```{r, fig.width=8, fig.height=10}

FM_EG_com_genes_UMAP$plot_merged_UMAP / FM_EG_com_genes_UMAP$plot_sep_UMAP

```

```{r, fig.width = 24, fig.height = 30}

FeaturePlot(object = FM_EG_com_genes_UMAP$merge_sub, features = common_list, ncol = 6)

```

<br>

## Comparison of all 

```{r, fig.width = 12, fig.height = 12}

(all_genes_UMAP$plot_merged_UMAP | 
   FM_genes_UMAP$plot_merged_UMAP | 
   FM_only_genes_UMAP$plot_merged_UMAP) / 
  (EG_genes_UMAP$plot_merged_UMAP | 
     EG_only_genes_UMAP$plot_merged_UMAP | 
     FM_EG_com_genes_UMAP$plot_merged_UMAP)

```


<div class = "red">
No visible differences on the UMAP whan we change the list of genes used to plot it. 
</div>


<br><br><br>


# Histogram of mRNA number => see QC graphs in classical analysis

## mRNA number per gene per cell

## mRNA number per gene per cell - cumulative sum 

## mRNAnumber 


<div class = "red">
Comments
</div>

<br><br><br>


# Conclusion 



<br><br><br>

****

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r}

# Clean working space and memory 
rm(list = ls())
gc()

```
