---
title: "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes"
author: "Laëtitia Racine"
date: "2023-03-20"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
div.beige {background-color:#FFE0B2; border-radius: 5px; padding: 8px;}
div.red {background-color:#FF8A80; border-radius: 5px; padding: 8px;}
div.orange {background-color:#FFB74D; border-radius: 5px; padding: 8px;}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(stringr)
library(tidyverse)
library(kableExtra)
library(ggVennDiagram)
library(ggplot2)

```

```{r, Working directories and constants}

start_time = Sys.time()
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/")
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/"))
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_ListGenes/", current_date, "/")

# Define a color code for list category 
color_code = c(
  "common_FM_EG" = "#6922FB",
  "only_EG" = "#F73131",
  "only_FM" = "#2E9EFF"
)

# Load external script with functions and constants
source(file = paste0(directory, "bin/Epi_FM_functions_constants.R"))

```


****

<br><br><br>



# Overview

<br>

Add comments here explaining what this code is aims at.




<br><br><br>



# Comparison of the list of genes 

<br>

**Load list of genes from EpiGift and FindMarkers**
```{r}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_FindMarkers_ListGenes/"))

## CTRL vs DON
genes_FM_list_CTRLDON = read.csv2(paste0(dir, "/", "CTRL-DON_list_219_genes_FindMarkers_th0.15_wilcox_pval0.05.csv"))
genes_FM_list_CTRLDON = unlist(as.vector(genes_FM_list_CTRLDON))
genes_EG_list_CTRLDON = read.csv2(paste0(directory_data, "20230309_EpiGift_Variable174Genes_pSNP5gt40.csv"), sep =",")
genes_EG_list_CTRLDON = genes_EG_list_CTRLDON$GeneName

# ## CTRL vs 2DG
# genes_FM_list_CTRL2DG = read.csv2(paste0(dir, "/", "CTRL-2DG_list_1021_genes_FindMarkers_th0.15_wilcox_pval0.05.csv"))
# genes_FM_list_CTRL2DG = unlist(as.vector(genes_FM_list_CTRL2DG))
# 
# ## CTRL vs AOA
# genes_FM_list_CTRLAOA = read.csv2(paste0(dir, "/", "CTRL-AOA_list_2569_genes_FindMarkers_th0.15_wilcox_pval0.05.csv"))
# genes_FM_list_CTRLAOA = unlist(as.vector(genes_FM_list_CTRLAOA))

```

<br>

**Compare lists of genes**
```{r}

## CTRL vs DON
common_list_CTRLDON = intersect(genes_FM_list_CTRLDON, genes_EG_list_CTRLDON)
only_FM_CTRLDON = setdiff(genes_FM_list_CTRLDON, genes_EG_list_CTRLDON)
only_EG_CTRLDON = setdiff(genes_EG_list_CTRLDON, genes_FM_list_CTRLDON)

cat("CTRL versus DON : \n",
    "FindMarkers - Seurat :",  length(genes_FM_list_CTRLDON), "genes \n",
    "EpiGIFT :", length(genes_EG_list_CTRLDON), "genes \n",
    "Common genes :", length(common_list_CTRLDON), "\n",
    "Only FindMarkers :", length(only_FM_CTRLDON), "\n",
    "Only EpiGift :", length(only_EG_CTRLDON), "\n")

saveRDS(object = common_list_CTRLDON, file = paste0(directory_output, "list_genes_CTRL-DON_commonEGFM.rds"))
saveRDS(object = only_FM_CTRLDON, file = paste0(directory_output, "list_genes_CTRL-DON_onlyFM.rds"))
saveRDS(object = only_EG_CTRLDON, file = paste0(directory_output, "list_genes_CTRL-DON_onlyEG.rds"))

```

<br>

**Represent a venn diagramm**
```{r}

## CTRL vs DON
ggVennDiagram(x = list(FindMarkers_CTRL_DON = genes_FM_list_CTRLDON, 
                       EpiGift_CTRL_DON = genes_EG_list_CTRLDON))+
  theme(legend.position = "none")


```



<br><br><br>



# Teta path on list only FindMarkers









<br><br><br>



# Log2FoldChange and Wilcoxon Test on list only EpiGift

```{r, Function definition FM manual}

FindMarkers_FC_wilcox = function(list_seurat, drugs_choice, list_genes) {
  
  # Select drugs and merge 
  merge_obj = merge(x = list_seurat[[drugs_choice[1]]], 
                    y = list_seurat[[drugs_choice[2]]],
                    add.cell.ids = drugs_choice,
                    merge.data = TRUE)
  
  # Variable features
  VariableFeatures(merge_obj) = rownames(merge_obj[["SCT"]]@scale.data)
  
  # Linear reduction and clustering
  merge_obj = RunPCA(object = merge_obj, 
                     assay = "SCT", 
                     npcs = 100,
                     reduction.name = "pca", 
                     seed.use = 42, 
                     verbose = FALSE)
  merge_obj = RunUMAP(object = merge_obj, 
                      assay = "SCT", 
                      dims = c(1:40), 
                      set.seed = 42,
                      reduction =  "pca", 
                      reduction.name = "umap", 
                      verbose = FALSE)
  merge_obj = FindNeighbors(object = merge_obj, assay = "SCT", dims = c(1:40))
  merge_obj = FindClusters(object = merge_obj, resolution = 0.6, algorithm = 3)
  
  # Prepare the object for FindMarkers
  Idents(merge_obj) <- "orig.ident"
  merge_obj <- PrepSCTFindMarkers(merge_obj)

  # Extract Matrix with cells from ident.1 and another with cells from ident.2
  ident1 = data.frame(merge_obj@assays$SCT@data) %>% dplyr::select(starts_with(paste0(drugs_choice[1], "_")))
  ident2 =  data.frame(merge_obj@assays$SCT@data) %>% dplyr::select(starts_with(paste0(drugs_choice[2], "_")))

  # Filter to keep only genes in list_genes
  ident1_f = ident1 %>% dplyr::filter(rownames(ident1) %in% list_genes)
  ident2_f = ident2 %>% dplyr::filter(rownames(ident2) %in% list_genes)
    
  # Compute expm1 on each value (exp(value) - 1)
  ident1_expm1 = expm1(ident1_f)
  ident2_expm1 = expm1(ident2_f)
  
  # Compute the mean +1 for each ident
  ident1_mean = rowMeans(ident1_expm1) + 1
  ident2_mean = rowMeans(ident2_expm1) + 1
  
  # Compute the log and the difference 
  ident1_log2 = log2(ident1_mean)
  ident2_log2 = log2(ident2_mean)
  avg_log2FC = ident1_log2 - ident2_log2
  df_avg_log2FC = data.frame(avg_log2FC) %>% tibble::rownames_to_column("gene")
  
  # Add a column to say if those genes would have been kept with the threshold value
  df_avg_log2FC = df_avg_log2FC %>% dplyr::mutate(pass_logfc.threshold = ifelse(abs(avg_log2FC) > 0.15, TRUE, FALSE))

  # Wilcoxon test
  tab_wilcox = list()
  for (i in 1:nrow(ident1_f)) {
    gene = rownames(ident1_f)[i]
    x = unlist(as.vector(ident1_f %>% dplyr::filter(rownames(ident1_f) == gene)))
    y = unlist(as.vector(ident2_f %>% dplyr::filter(rownames(ident2_f) == gene)))
    tab_wilcox[gene] = wilcox.test(x, y)$p.value
  }
  tab_wilcox = as.data.frame(do.call(rbind, tab_wilcox)) 
  colnames(tab_wilcox) = "p_val"
  tab_wilcox = tab_wilcox %>% 
    tibble::rownames_to_column("gene") %>%
    # manual bonferroni correction (multiply the pvalue by the total number of comparison (all genes of the seurat obj))
    # set the biggest value at 1 (same as if we had used p.adjust(tab_wilcox$p_val, "bonferroni", n=nrow(merge_obj)))
    dplyr::mutate(p_val_adj = tab_wilcox$p_val * nrow(merge_obj)) %>%
    dplyr::mutate(p_val_adj = ifelse(p_val_adj<1, p_val_adj, 1))
  
  # Join table with avg_log2FC
  final_tab = left_join(tab_wilcox, df_avg_log2FC, by = "gene")
  
  # output
  return(final_tab)
  
}

```

```{r, Loading seurat obj}

list_seurat_obj = readRDS(paste0(directory_data, "list_seurat_obj_epi.rds"))

list_seurat_obj = lapply(list_seurat_obj, function(x) { 
    SCTransform(object = x, 
                assay = "RNA", 
                new.assay.name = "SCT",
                do.center = TRUE, 
                do.scale = FALSE,
                variable.features.n = NULL,
                variable.features.rv.th = 1.3,
                residual.features = NULL,
                vars.to.regress = NULL,
                min_cells = 1,
                ncells = NULL,
                verbose = FALSE,
                # vst additional parameter
                method = "glmGamPoi",
                residual_type = "deviance",
                n_genes = NULL)
  })

saveRDS(list_seurat_obj, paste0(directory_output, "list_seurat_obj_epi_sct.rds"))

```

```{r, Call function FM manual}

EG_FindMarkerscalcul_CTRLDON = FindMarkers_FC_wilcox(list_seurat = list_seurat_obj,
                                                     drugs_choice = c("CTRL", "DON"), 
                                                     list_genes = genes_EG_list_CTRLDON)

# EG_FindMarkerscalcul_CTRL2DG= FindMarkers_FC_wilcox(list_seurat = list_seurat_obj,
#                                                     drugs_choice = c("CTRL", "2DG"), 
#                                                     list_genes = genes_EG_list_CTRL2DG)
# 
# EG_FindMarkerscalcul_CTRLAOA = FindMarkers_FC_wilcox(list_seurat = list_seurat_obj,
#                                                      drugs_choice = c("CTRL", "AOA"), 
#                                                      list_genes = genes_EG_list_CTRLAOA)

saveRDS(EG_FindMarkerscalcul_CTRLDON, 
        paste0(directory_output, current_date, "_EG_FindMarkerscalcul_CTRLDON.rds"))
# saveRDS(EG_FindMarkerscalcul_CTRL2DG,
#         paste0(directory_output, current_date, "_EG_FindMarkerscalcul_CTRL2DG.rds"))
# saveRDS(EG_FindMarkerscalcul_CTRLAOA,
#         paste0(directory_output, current_date, "_EG_FindMarkerscalcul_CTRLAOA.rds"))

```

```{r, Visualize the result FM manual, fig.width = 16, fig.height = 6}

tab_CTRLDON = EG_FindMarkerscalcul_CTRLDON
# tab_CTRL2DG = EG_FindMarkerscalculCTRL2DG
# tab_CTRLAOA = EG_FindMarkerscalculCTRLAOA

tab_CTRLDON %>%
  kable(caption = "CTRL vs DON : pvalue for genes from EpiGift that passed the log2fc threshold") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")


plot_FC = ggplot(tab_CTRLDON, aes(x = gene, y = abs(avg_log2FC))) +
  geom_point() +
  geom_hline(yintercept = 0.15, linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  ggtitle("CTRL vs DON - EpiGIFT list - Threshold on avg_log2FC value")
plot_FC

cat("Number of genes that passed the threshold on log fold change :")
table(tab_CTRLDON$pass_logfc.threshold)

plot_pval = ggplot(tab_CTRLDON, aes(x = gene, y = p.adj)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  ggtitle("CTRL vs DON - EpiGIFT list - pvalue adjusted of all EpiGIFT genes")
plot_pval

plot_pval_cut = ggplot(tab_CTRLDON %>% dplyr::filter(p.adj < 0.25), aes(x = gene, y = p.adj)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  ggtitle("CTRL vs DON - EpiGIFT list - pvalue adjusted of all EpiGIFT genes")
plot_pval_cut

plot_pval_filter = ggplot(tab_CTRLDON %>% dplyr::filter(pass_logfc.threshold  == TRUE), 
                   aes(x = gene, y = p.adj)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  ggtitle("CTRL vs DON - EpiGIFT list - pvalue adjusted of genes that passed the log2FC threshold")
plot_pval_filter

genes_EG_passFM_CTRLDON = (tab_CTRLDON %>% dplyr::filter(pass_logfc.threshold  == TRUE))$gene
setdiff(genes_EG_passFM_CTRLDON, common_list_CTRLDON)

cat("We didn't identify all the genes from EpiGIFT method with the FindMarkers method because they didn't pass the log2FC threshold.")

```

<br>




# Conclusion

Add conclusion here !






<br><br><br>


****

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results='hide'}

# Clean working space and memory 
rm(list = ls())
gc()

```