---
title: "scRNAseq_EpiGIFT_Seurat_InvestigationCompare_Random"
author: "Laëtitia Racine"
date: "2023-02-15"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---
    
```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(sctransform)
library(stringr)
library(tidyverse)
library(kableExtra)

```

```{r, Working directories and external script}

start_time = Sys.time()
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/")
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_Random/"))
dir.create(path = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_Random/"), current_date)
directory_output = paste0(directory, "exp/scRNAseq_EpiGIFT_Seurat_InvestigationCompare_Random/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/Epi_FM_functions_constants.R"))

```


****


<br><br><br>



# Overview 

<br>

We have lists from FindMarkers with different thresholds and Wilcoxon test (cf. scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat).      
We have two lists from EpiGift depending the number of bin fixed (bin 70 => 130 genes, bin 30 => 365 genes).  
We have a certain number of common genes between Epigift lists and FM lists.    
We want to see :    
1) if the overlapping number increase when we change FM threshold on logFC    
2) if this number can be just "luck" so we compare to a random selection of genes    

```{r, Input loading}

# List of 130 genes with threshold bin 70 in EpiGift
epi_130 = read.csv2(paste0(directory, "data/", "20230220_EpiGift_Variable130Genes.csv"), sep = ",")
epi_130 = epi_130[,2]
# List of 365 genes with threshold bin 30 in EpiGift
epi_365 = read.csv2(paste0(directory, "data/", "20230220_EpiGift_Variable365Genes.csv"), sep = ",")
epi_365 = epi_365[,2]
# List of FM genes and object merge_fm
dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_EpiGIFT_Seurat_InvestigationGenesSeurat/"))
list_FM_genes = readRDS(paste0(dir, "/list_genes_FM_different_thresholds.rds"))
merge_fm = readRDS(paste0(dir, "/merge_fm.rds"))

```
  
  
<br><br><br>
  
  
# FM threshold and random gene selection  
  
```{r}

# See if the overlap increases proportionally or not

list_overlap_FM_epi130 = list()
list_overlap_FM_epi365 = list()
list_overlap_FM_random130 = list()
list_overlap_FM_random365 = list()

for (i in 1:length(list_FM_genes)) {
  
  threshold = names(list_FM_genes)[i]
  
  # List of 130 and 365 random genes extract from the matrix to compare
  random_130 = sample(rownames(merge_fm), 130)
  random_365 = sample(rownames(merge_fm), 365)
  
  # List of FM genes
  genes = list_FM_genes[[i]]
  
  list_overlap_FM_epi130[threshold] = list(intersect(epi_130, genes))
  list_overlap_FM_epi365[threshold] = list(intersect(epi_365, genes))
  list_overlap_FM_random130[threshold] = list(intersect(random_130, genes))
  list_overlap_FM_random365[threshold] = list(intersect(random_365, genes))
  
}

tab_comp = data.frame(Nbgenes_FindMarkers = lengths(list_FM_genes),
                      Nbgenes_overlap_FM_epi130 = lengths(list_overlap_FM_epi130),
                      Nbgenes_overlap_FM_epi365 = lengths(list_overlap_FM_epi365),
                      Nbgenes_overlap_FM_random130 = lengths(list_overlap_FM_random130),
                      Nbgenes_overlap_FM_random365 = lengths(list_overlap_FM_random365)) %>%
  tibble::rownames_to_column("FindMarkers_logfc_threshold")

tab_comp %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

tab_plot = tab_comp %>% 
  tidyr::pivot_longer(cols=c(2:ncol(tab_comp)),
                      names_to="Category",
                      values_to="Nbgenes")

plot = ggplot(tab_plot, aes(x=FindMarkers_logfc_threshold, y=Nbgenes, 
                            shape=Category, col=Category)) +
  geom_point()
plot

ggsave(plot = plot, filename = paste0(directory_output, "overlap_EpiGiftgenes_FindMarkersgenes.png")) 
write.csv2(x = tab_comp, file = paste0(directory_output, "tab_comp_EpiGift_FindMarkers.csv"))

```


<br><br><br>


# Conclusion 

When we increase the threshold on logfc for FindMarkers function :  
- the number of significant genes in FindMarkers list decreases  
- the number of genes overlapping with EpiGIFT genes (130 and 365) decreases also BUT not propotionnaly  
When we compared our FM list with the same number of random genes than EpiGIFT list, we have much less common genes => the overlap between EpiGIFT and FM list is not just random luck.  
We keep logfc.threshold = 0.15 for FindMarkers function.  


<br><br><br>


****

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results='hide'}

# Clean working space and memory 
rm(list = ls())
gc()

```
